---
title: "Project 1 Final Deliverable"
author: "Hailey Ernest and Nicole Yee"
format: html
editor: visual
embed-resources: true
theme: sandstone
code-fold: true
---

## Data Context

```{r}
#| label: Load in Data and libraries
#| results: 'hide'
#| message: false
#| warning: false
#| echo: false

library(tidyverse)
library(here)
library(readxl)
library(janitor)
library(gt)
library(xfun)
library(httr)
library(tidyjson)
library(jsonlite)
library(rvest)
library(chromote)
library(hexbin)
library(ggridges)
library(viridis)
library(patchwork)
library(glue)
library(stringr)
library(stringi)
library(patchwork)

cases_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv') |>
  clean_names()
cases_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')

gdp_web <- readr::read_csv('gdp_web.csv')
```

```{r}
#| label: Webscraping GDP data
#| warning: false
#| message: false
#| eval: false

# function to scrape one year of gdp data
scrape_gdp_year <- function(year) {
  
  # create url for a specific year
  url <- glue("https://ourworldindata.org/grapher/gdp-per-capita-worldbank?tab=table&time={year}..{year}&tableFilter=countries")
  
  # open browser session: page is dynamic (live)
  b <- ChromoteSession$new()
  
  # browser closes after function
  on.exit(try(b$close(), silent = TRUE), add = TRUE)
  
  # go to page
  b$Page$navigate(url)
  
  # wait for loaded table
  Sys.sleep(5)
  
  # get rendered html from the page
  eval_result <- b$Runtime$evaluate("document.documentElement.outerHTML")
  rendered_html <- eval_result$result$value
  
  # read html into r
  page <- read_html(rendered_html)
  
  # extract the table from the page
  tab <- page |>
    html_element("table") |>
    html_table()
  
  # keep only country and gdp
  tab_clean <- tab[, 1:2]
  names(tab_clean) <- c("country", "gdp_per_capita")
  
  # clean the data
  tab_clean <- tab_clean |>
    mutate(
      scraped_year = year, # year                              
      gdp_per_capita = str_remove_all(gdp_per_capita, "\\$|,"), # remove $ and ,
      gdp_per_capita = na_if(gdp_per_capita, ""), # empty is NA
      gdp_per_capita = as.numeric(gdp_per_capita) # convert to numeric
    ) |>
    filter(country != "Country or region") |>
    filter(!is.na(country))
  tab_clean
}

# create vector of years to scrape
years <- 2012:2024

# loop over all years, then combine
gdp_web <- map_dfr(years, scrape_gdp_year)
```

```{r}
#| label: GDP data joining and preperation
#| warning: false
#| message: false

# Ensure countries are joined cleanly
clean_country <- function(x) {
  x |>
    stri_trans_general("Latin-ASCII") |>
    str_to_lower() |>
    str_replace_all("&", "and") |>
    str_remove_all("[[:punct:]]") |>
    str_squish()
}

# Change country names that don't align with the measles data
gdp_web_clean <- gdp_web |>
  rename(year = scraped_year) |>
  mutate(
    year = as.numeric(year),
    country_join = recode(country,
      "United States" = "United States of America",
      "Bolivia" = "Bolivia (Plurinational State of)",
      "Venezuela" = "Venezuela (Bolivarian Republic of)"
    ),
    country_key = clean_country(country_join)
  )

# Clean countries and years
cases_year_clean <- cases_year |>
  mutate(
    year = as.numeric(year),
    country_key = clean_country(country)
  )

# join the yearly gdp with yearly cases data
cases_year_gdp <- cases_year_clean |>
  left_join(
    gdp_web_clean |>
      select(country_key, year, gdp_per_capita),
    by = c("country_key", "year")
  )

# join the yearly gdp/cases data to the monthly cases data
cases_gdp <- cases_month |>
  left_join(
    cases_year_gdp,
    by = c("country", "year"),
    suffix = c("_month", "_year")
  ) |>
  # make a tibble to join
  as_tibble()

# create yearly country-level dataset with GDP joined, and clean readable names
cases_gdp_year <- cases_year |>
  filter(year != 2025) |>
  mutate(
    year = as.numeric(year),
    measles_epi_linked = as.numeric(measles_epi_linked)
  ) |>
  left_join(
    gdp_web |>
      rename(year = scraped_year) |>
      select(country, year, gdp_per_capita),
    by = c("country", "year")
  ) |>
  mutate(
    region = recode(region,
      "AFRO" = "Africa",
      "AMRO" = "Americas",
      "EMRO" = "Eastern Mediterranean",
      "EURO" = "Europe",
      "SEARO" = "South-East Asia",
      "WPRO" = "Western Pacific"
    )
  ) |>
  filter(
    !is.na(measles_epi_linked),
    !is.na(gdp_per_capita),
    measles_epi_linked > 0
  )
```

We utilized data from several sources. Our report is mainly centered around measles and rubella cases, but we incorporated additional data about GDP.

Our main set of data presents suspected and confirmed classification of measles and rubella cases across regions and countries in the world. It is presented in two forms: monthly case counts (cases_month) and yearly summaries (cases_year). The data is collected and reported by healthcare providers and national surveillance systems and compiled by the World Health Organization (WHO). The data was downloaded on June 12, 2025, and is finalized by July of each year. The data covers regions and countries worldwide. The data is collected to track and confirm measles and rubella cases across populations over time. Suspected measles cases are identified based on symptoms or clinical judgment, reported through national surveillance systems to the WHO, and then classified as clinically compatible, epidemiologically linked, or laboratory-confirmed, with total measles cases calculated as the sum of these categories for most countries.

We additionally used country-level GDP per capita data, where the observations are countries and regions around the world. The data includes over 200 countries' GDP, and ranges from 1990 to 2024. GDP per capita is calculated by dividing a country’s total output by its population and adjusting values with the same international dollars to take into account inflation and living cost differences. The goal of this data is to compare economic performance, living conditions, and economic growth across almost all countries over many years. The data is created by organizations such as the World Bank, IMF, OECD, and Eurostat, and is summarized by Our World in Data. GDP data is collected by national statistical agencies using censuses, surveys, and tax or government data, and then analyzed and organized by the World Bank, IMF, etc.

## Research Questions

1.  Globally, has there been improvement in containing measles and rubella cases in recent years and is there a time of year where these cases become more common?
2.  Are countries with lower GDP per capita associated with higher numbers of epidemiologically linked measles cases and does this relationship vary by region?

For the remainder of this report, we will be referencing research question 1 as "Measles and Rubella Cases Over Time" and research question 2 as "Measles Association with GDP per capita".

## Data Summaries & Visualizations

### Research Question 1: Measles and Rubella Cases Over Time

```{r}
#| fig-cap-location: bottom
#| fig-cap: Figure 1. Trends in Measles and Rubella Cases
#| fig-alt: "Line plot showing total global measles and rubella cases by year from 2012 to 2024, with measles showing larger spikes and greater variability than rubella."

annotate_text <- tibble(
  year = c(2014.5, 2014.3),
  y_lab = c(75000, 200000),
  disease = c("Rubella", "Measles")
)

cases_year |>
  group_by(year) |>
  summarize(measles = sum(measles_total, na.rm = TRUE),
            rubella = sum(rubella_total, na.rm = TRUE)) |>
  ggplot() +
  geom_ribbon(aes(x = year, ymin = measles, ymax = rubella),
              position = "identity",
              inherit.aes = FALSE,
              fill = "#F8F5F0") +
  geom_line(aes(x = year,
                y = measles,
                color = "Measles"), 
            linewidth = 1) +
  geom_line(aes(x = year,
                y = rubella,
            color = "Rubella"),
            linewidth = 1) +
  scale_color_manual(name = "Disease",
                     values = c("Measles" = "#325D88", "Rubella" = "#3E3F3A")) +
  scale_x_continuous(breaks = unique(cases_year$year),
                     expand = c(0.01,0.01)) +
  scale_y_continuous(breaks = seq(0, 600000, by = 100000),
                     labels = scales::label_comma()) +
  geom_text(data = annotate_text,
            mapping = aes(x = year,
                          y = y_lab,
                          label = disease,
                          color = disease),
            inherit.aes = FALSE,
            show.legend = FALSE
            ) +
  theme_bw() +
  labs(x = "Year",
       y = "Number of Cases",
       title = "Measles and Rubella Cases Over Time",
       subtitle = "Similar spikes in number of cases occur between diseases, but measles \ncases are more volatile in the magnitude of change.") +
  theme(panel.grid.minor = element_blank())
```

Surprisingly, as shown in Figure 1, there is not a consistent linear trend in global cases for either disease. It does appear that there is a slight generally downward trend in rubella cases, but there still are small spikes in 2013, 2019, and 2024. Total measles cases appears to have the same peaks, but at a much larger magnitude. Recent years have had decreasing number of cases, but overall, there is not a noticeable dip in measles cases.

```{r}
#| tbl-cap-location: bottom
#| tbl-cap: Table 1. Case Counts Compared to the Median Case Count

cases_year |>
  group_by(year) |>
  summarize(
    total_measles = sum(measles_total, na.rm = TRUE),
    total_rubella = sum(rubella_total, na.rm = TRUE),
    .groups = "drop") |>
  mutate(
    median_measles = median(total_measles),
    median_rubella = median(total_rubella),
    measles_vs_median = total_measles - median_measles,
    rubella_vs_median = total_rubella - median_rubella,
    measles_pct = (total_measles - median_measles) / median_measles,
    rubella_pct = (total_rubella - median_rubella) / median_rubella) |>
  select(year, measles_pct, rubella_pct) |>
  gt()  |>
  cols_label(
     year = md("**Year**"),
     measles_pct = md("**Measles**"),
     rubella_pct = md("**Rubella**")
   ) |>
   # add a table title
   tab_header(title = md("Difference in Measles and Rubella Cases Each Year Compared to the Typical Number of Cases")) |>
   # add a table caption
   tab_caption(caption = md("Percent change in cases compared to the overall median number of measles and rubella cases.")) |>
   # bold column labels
   tab_style(
     style = cell_text(weight = "bold"),
     locations = cells_column_labels()
   ) |>
  fmt_percent(columns = c(measles_pct, rubella_pct), decimals = 2) 
```

Percent differences from the median case count of each disease supports our findings in Figure 1. In the peak years 2013, 2019, and 2024, there appears to be large differences from the overall median number is cases for both diseases. The largest difference from the median was in 2019, with the second largest difference in 2024 (Table 1). Interestingly enough, a majority of the recent years aside from 2024 actually had less total cases than the median. The same can be said for rubella, which has mostly had less total cases than the median.

```{r}
#| fig-cap-location: bottom
#| fig-cap: Figure 2. Measles and Rubella Cases by Month
#| warning: false
#| fig-alt: "Ridgeline density plot of the distribution of global measles case counts across months, showing variation in case frequency throughout the year."

color_gradient <- colorRampPalette(c("#A6DAF5", "#325D88"))(12)
names(color_gradient) <- month.name

measles <- cases_month |>
  group_by(month, year) |>
  summarise(measles_total = sum(measles_total, na.rm = TRUE)) |>
  mutate(month_full = month.name[month]) |>
  ggplot(aes(x = measles_total,
             y = reorder(month_full, -month),
             group = month,
             fill = month_full)) +
  geom_density_ridges() +
  # geom_density_ridges_gradient(scale = 2, rel_min_height = 0.00000000001,
  #                              alpha = 0.5) +
  # scale_fill_viridis(name = "Number of Cases", option = "C") +
  scale_x_continuous(breaks = seq(0, 140000, by = 40000),
                     labels = scales::label_comma()) +
  #AGDAF5  #325D88 
  scale_fill_manual(values = color_gradient) +
  scale_y_discrete(expand = c(0.0001,0.5)) +
  theme_minimal() +
  labs(x = "Number of Cases",
       y = "",
       title = "Measles") +
  theme(panel.grid.minor.x = element_blank(),
        legend.position = "none")


color_gradient <- colorRampPalette(c("#B8B9B6", "#3E3F3A"))(12)
names(color_gradient) <- month.name

rubella <- cases_month |>
  group_by(month, year) |>
  summarise(rubella_total = sum(rubella_total, na.rm = TRUE)) |>
  mutate(month_full = month.name[month]) |>
  ggplot(aes(x = rubella_total,
             y = reorder(month_full, -month),
             group = month,
             fill = month_full)) +
  geom_density_ridges() +
  scale_x_continuous(breaks = seq(0, 20000, by = 4000),
                     labels = scales::label_comma()) +
  scale_fill_manual(values = color_gradient) +
  scale_y_discrete(expand = c(0.0001,0.5)) +
  theme_minimal() +
  labs(x = "Number of Cases",
       y = "",
       title = "Rubella") +
  theme(panel.grid.minor.x = element_blank(),
        legend.position = "none")


(measles + rubella) +
  plot_layout(guides = "collect",
              axis_titles = "collect") +
  plot_annotation(title = "Comparison of Global Measles and Rubella Cases for Each Month",
                  subtitle = "Monthly distributions span case counts from 2012 to 2025")
```

Measles and rubella are highly contagious diseases, but it appears they slightly differ in the time of year they are most common. More measles cases seem to occur during late winter months, with more cases being observed starting November or December, but noticeably increases January to May (Figure 2). Higher case counts of rubella appear similarly in overlapping months, but the period of common cases is shifted forward, from January to June or July (Figure 2).

```{r}
#| tbl-cap-location: bottom
#| tbl-cap: Table 2. Months with the Most Measles Cases

# function to get the month of a year with the highest number of cases
get_max_month <- function(yr){
  cases_month |>
    filter(year == yr) |>
    group_by(month) |>
    summarize(num_measles = sum(measles_total, na.rm = TRUE),
              num_rubella = sum(rubella_total, na.rm = TRUE),
              .groups = "drop") |>
    summarize(
      measles_month = month[which.max(num_measles)],
      measles_cases = max(num_measles),
      rubella_month = month[which.max(num_rubella)],
      rubella_cases = max(num_rubella)
    ) |>
    mutate(year = yr, .before = 1)
}

cases_year |>
  distinct(year) |>
  pull(year) |>
  # apply function to each year in the data
  map_dfr(get_max_month) |>
  # convert month abbreviations to month names
  mutate(measles_month = month.name[measles_month],
         rubella_month = month.name[rubella_month]) |>
  # # make a gt table object
  gt() |>
  # rename and bold each column
  cols_label(
     year = md("**Year**"),
     measles_month = md("**Month**"),
     measles_cases = md("**Cases**"),
     rubella_month = md("**Month**"),
     rubella_cases = md("**Cases**")
   ) |>
   # add a table title
   tab_header(title = md("Time of Year with Increased Measles and Rubella Cases Over Time")) |>
   # add a table caption
   tab_caption(caption = md("Months with the Highest Number of Cases.")) |>
   # bold column labels
   tab_style(
     style = cell_text(weight = "bold"),
     locations = cells_column_labels()
   ) |>
    tab_spanner(
      label = "Measles",
      columns = c(measles_month, measles_cases)
    ) |>
    tab_spanner(
      label = "Rubella",
      columns = c(rubella_month, rubella_cases)
    ) |>
   fmt_number(columns = c(3,5), sep_mark = ",", decimals = 0)
```

We further investigated the observed global case counts to find the month with the most cases of each disease over time. As shown in Table 2, the month with the highest frequency of having the most measles cases in a year was March, with February following. March is also the month that has had the most rubella cases, and April as the second most frequent month (Table 2). These summaries similarly support the findings of Figure 2.

### Research Question 2: Measles Association with GDP per capita

```{r}
#| fig-cap-location: bottom
#| fig-cap: Figure 3. Epidemiologically Linked Measles Cases and GDP per Capita by Region.
#| message: false
#| fig-width: 16
#| fig-height: 12
#| fig-alt: "Faceted scatterplot of epidemiologically linked measles cases over time by region, with log-scaled y-axis and point color representing GDP per capita."

# faceted plot by region
ggplot(cases_gdp_year,
       aes(x = year,
           y = measles_epi_linked,
           color = gdp_per_capita)) +
  geom_jitter(alpha = 0.9, size = 3, width = 0.25, height = 0.1) +
  scale_y_log10(labels = scales::comma) +
  scale_x_continuous(breaks = c(2012, 2016, 2020, 2024)) +
  scale_color_gradient(
    low = "#A6DAF5",
    high = "#3E3F3A",
    labels = scales::comma
  ) +
  facet_wrap(~ region, ncol = 2, scales = "free_y") +
  labs(
    title = "Epidemiologically Linked Measles Cases Over Time by Region",
    subtitle = "Each point represents a country in a year & color shows GDP per capita",
    x = "Year",
    y = "Epidemiologically linked measles cases (log scale)",
    color = "GDP per capita"
  ) +
  theme_minimal(base_size = 25)
```

As shown in Figure 3, there are regional differences in epidemiologically linked measles cases for changes GDP per capita. Africa and South-East Asia show a range of case counts for countries, but are mostly lower GDP countries. The Americas have very few epidemiologically linked cases across years, suggesting that cases in this region are usually linked less through geographic transmission patterns. Europe has varying case counts across years and don't appear to be associated with GDP. However, the Eastern Mediterranean and Western Pacific regions show a clear pattern that higher GDP countries tend to have lower numbers of epidemiologically linked measles cases, and lower GDP countries tend to have higher case counts.

```{r}
#| tbl-cap-location: bottom
#| tbl-cap: Table 3. Confirmed Measles Cases that are Epidemiologically-linked

# function to get proportion of epidemiologically-linked cases for each year in one region
epi_proportions <- function(region_code) {
  cases_month |>
    filter(.data$region == region_code) |>
    group_by(year) |>
    # get total counts for each year
    summarise(
      # measles total epidemiologically-linked and lab confirmed
      measles_epi_linked = sum(measles_epi_linked, na.rm = TRUE),
      measles_lab_confirmed = sum(measles_lab_confirmed, na.rm = TRUE),
      .groups = "drop"
    ) |>
    # modify columns by calculating proportions
    mutate(
      prop_epi_measles = measles_epi_linked / 
        (measles_epi_linked + measles_lab_confirmed)
    ) |>
    select(year, prop_epi_measles)
}

cases_month |>
  distinct(region) |>
  # rename regions to full names
  mutate(
    region_name = recode(region,
      "AFR" = "Africa",
      "AMR" = "Americas",
      "EMR" = "Eastern Mediterranean",
      "EUR" = "Europe",
      "SEAR" = "South-East Asia",
      "WPR" = "Western Pacific"
    )
  ) |>
  # apply function to get proportions for each region
  mutate(epi_props = map(region, epi_proportions)) |>
  # get columns from returned map lists
  unnest(epi_props) |>
  # keep only region name, year, and proportion
  select(region_name, year, prop_epi_measles) |>
  # group by region so the table is divided by region
  group_by(region_name) |>
  # make a gt table
  gt() |>
  # format the proportion columns to be percents 
  fmt_percent(columns = matches("prop"), decimals = 1) |>
  # bold region group labels
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) |>
  # rename and bold column names
  cols_label(
    region_name = md("**Region**"),
    year = md("**Year**"),
    prop_epi_measles = md("**Epidemiologically-linked measles cases**")
  ) |>
  # add a table title
  tab_header(
    title = md("Proportion of Confirmed Cases that are Epidemiologically-linked")
  ) |>
  # add a table caption
  tab_caption(
    caption = md("Changes in each region's confirmed measles cases makeup over time.")
  )
```

In Table 3, the proportion of confirmed measles cases that are epidemiologically linked continue to show variation across regions and over time. Africa and South-East Asia have higher proportions, meaning a larger number of cases are linked through geographic transmission. Europe and the Eastern Mediterranean show moderate and slightly varying proportions over time. The Americas and Western Pacific regions show very low proportions over time, and that less cases are epidemiologically linked.

```{r}
#| tbl-cap-location: bottom
#| tbl-cap: Table 4. Epidemiologically-linked counts by GDP levels of regions.

# calculation of global GDP level groups: low, medium, and high
cases_gdp_year_global <- cases_gdp_year |>
  mutate(
    gdp_group = ntile(gdp_per_capita, 3),
    gdp_group = recode(as.character(gdp_group),
      "1" = "Low GDP",
      "2" = "Medium GDP",
      "3" = "High GDP"
    ),
    gdp_group = factor(gdp_group,
      levels = c("Low GDP", "Medium GDP", "High GDP")
    )
  )

# function to get total epidemiologically-linked cases by GDP group for one region
gdp_epi_groups <- function(region_name) {
  cases_gdp_year_global |>
    filter(.data$region == region_name) |>
    group_by(gdp_group) |>
    summarise(
      epi_linked_cases = sum(measles_epi_linked, na.rm = TRUE),
      .groups = "drop"
    )
}

cases_gdp_year |>
  distinct(region) |>
  # apply function to get GDP groups for each region
  mutate(gdp_epi = map(region, gdp_epi_groups)) |>
  # get columns from returned map lists
  unnest(gdp_epi) |>
  # keep GDP levels with 0 cases
  complete(region, gdp_group, fill = list(epi_linked_cases = 0)) |>
  # group by region so the table is divided by region
  group_by(region) |>
  # make a gt table
  gt() |>
  # format the epidemiologically-linked case counts
  fmt_number(columns = epi_linked_cases, decimals = 0) |>
  # bold region group labels
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) |>
  # rename and bold column names
  cols_label(
    region = md("**Region**"),
    gdp_group = md("**GDP Level**"),
    epi_linked_cases = md("**Epidemiologically-linked measles cases**")
  ) |>
  # add a table title
  tab_header(
    title = md("Epidemiologically-linked Measles Cases by Regional GDP Level")
  ) |>
  # add a table caption
  tab_caption(
    caption = md("Total epidemiologically-linked measles cases grouped by global GDP per capita
                 low, medium, and high within each region. Based on yearly country observations
                 from 2012–2024.")
  )
```

Total epidemiologically linked measles cases differ across GDP levels within regions. In Africa and the Eastern Mediterranean, most cases occur in low GDP countries which supports the trends in the visualizations of regions (Figure 3). South-East Asia and Western Pacific show a concentration of cases in medium GDP countries, with Western Pacific having few overall cases. The Americas have relatively very few cases overall. Europe differs from other regions by most cases occurring in medium and high GDP countries, showing the higher GDP trend across countries within the region (Table 4).

```{r}
#| fig-cap-location: bottom
#| fig-cap: "Figure 4. Epidemiologically Linked Measles Cases and GDP per Capita by Region"
#| warning: false
#| fig-alt: "Two side-by-side ridgeline density plots by region. The left plot shows the distribution of epidemiologically linked measles cases across country-year observations from 2012 to 2024. The right plot shows the distribution of GDP per capita across the same regions and years."
#| fig-width: 18
#| fig-height: 8

epi_region_plot <- cases_gdp_year |>
  ggplot(aes(x = log1p(measles_epi_linked),
             y = region,
             fill = ..x..)) +
  geom_density_ridges_gradient(scale = 2, rel_min_height = 0.01) +
  scale_x_continuous(
    breaks = log1p(c(0, 10, 100, 1000, 10000)),
    labels = scales::comma(c(0, 10, 100, 1000, 10000))
  ) +
  scale_fill_gradient(
    low = "#F8F5F0",
    high = "#325D88"
  ) +
  scale_y_discrete(expand = c(0.0001, 0.5)) +
  labs(
    #title = "Distribution of Epidemiologically Linked Measles Cases by Region",
    #subtitle = "Country-year observations from 2012–2024",
    x = "Epidemiologically linked measles cases (log scale)",
    y = "",
    fill = "Cases"
  ) +
  theme_minimal(base_size = 20) +
  theme(
    panel.grid.minor.x = element_blank(),
    legend.position = "none"
  )

gdp_region <- cases_gdp_year |>
  ggplot(aes(x = gdp_per_capita,
             y = region,
             fill = ..x..)) +
  geom_density_ridges_gradient(scale = 2, rel_min_height = 0.01) +
  scale_x_log10(labels = scales::label_comma()) +
  scale_fill_gradient(
    low = "#A6DAF5",
    high = "#3E3F3A"
  ) +
  scale_y_discrete(expand = c(0.0001, 0.5)) +
  labs(
    #title = "Distribution of GDP per Capita by Region",
    #subtitle = "Country-year observations from 2012–2024",
    x = "GDP per capita (log scale)",
    y = "",
    fill = "GDP per capita"
  ) +
  theme_minimal(base_size = 20) +
  theme(
    panel.grid.minor.x = element_blank(),
    legend.position = "none"
  )

epi_region_plot + gdp_region +
  plot_layout(ncol = 2) +
  plot_annotation(
    title = "Distributions of Measles Cases and GDP per Capita by Region",
    subtitle = "These plots show yearly country distributions from 2012–2024",
    theme = theme(
      plot.title = element_text(size = 26),
      plot.subtitle = element_text(size = 18)
    )
  )
```

To understand the findings of differences in epidemiologically linked measles cases and GDP per capita across regions, the distribution of each is visualized in Figure 4. Epidemiologically linked cases show the Americas have the lowest case levels, Europe, Eastern Mediterranean, and the Western Pacific have slightly higher cases, and Africa and South-East Asia have the highest concentrations of cases. The distribution of GDP per capita shows that Africa has the lowest GDP levels, then Eastern Mediterranean and South-East Asia, followed with the Americas and Western Pacific that have higher GDP levels, and Europe has the highest overall. This explains the patterns found in Figure 3, as limited GDP differences within Europe, Africa, and South-East Asia makes it challenging to identify a clear relationship between GDP and epidemiologically linked measles cases.

## Data Model

### Research Question

#### Are countries with lower GDP per capita associated with higher numbers of epidemiologically linked measles cases and does this relationship vary by region?

This model evaluates the relationship between GDP per capita and epidemiologically linked measles cases. GDP per capita is used as a measure of a country’s economic development, and epidemiologically linked cases show patterns of disease transmission based on geography. The model uses country-year observations from 2012 to 2024. The variables are both log-transformed to account for skewness across countries, since changes at lower income levels might have a larger impact than at higher income levels. This is consistent with used approaches in recent global health research.

Region is included as a categorical variable to account for structural differences in disease transmission and reporting across geographic areas, and an interaction between GDP and region is included to assess whether the relationship between GDP per capita and epidemiologically linked measles cases varies across regions.

```{r}
#| warning: false
#| tbl-cap-location: bottom
#| tbl-cap: Table 5. Model results analyzing GDP and region effects on epidemiologically linked measles cases.

# create model dataset
epi_gdp_model_data <- cases_gdp |>
  distinct(country, year, region_year, gdp_per_capita, measles_epi_linked_year) |>
  rename(region = region_year) |>
  filter(
    year != 2025,
    gdp_per_capita > 0
  ) |>
  drop_na(gdp_per_capita, measles_epi_linked_year, region) |>
  mutate(
    # make region a categorical variable
    region = factor(region),
    # log transform variables to reduce skew
    log_gdp = log10(gdp_per_capita),
    log_epi_linked = log1p(measles_epi_linked_year)
  )

# fit model with GDP, region, and GDP-by-region interaction
epi_gdp_model <- lm(
  log_epi_linked ~ log_gdp * region,
  data = epi_gdp_model_data
)

# make coefficient table from model summary
coef_table <- as.data.frame(summary(epi_gdp_model)$coefficients) |>
  rownames_to_column(var = "term")

coef_table |>
  gt() |>
  # format values
  fmt_number(
    columns = c(Estimate, `Std. Error`, `t value`, `Pr(>|t|)`),
    decimals = 3
  ) |>
  sub_small_vals(
    columns = `Pr(>|t|)`,
    threshold = 0.001
  ) |>
  cols_align(align = "center") |>
  # rename column headers
  cols_label(
    term = "Model Term",
    Estimate = "Estimate",
    `Std. Error` = "Standard Error",
    `t value` = "t Statistic",
    `Pr(>|t|)` = "p-value"
  ) |>
  # add title and subtitle
  tab_header(
    title = "Model Results: GDP, Region, and Epidemiologically Linked Measles Cases",
    subtitle = "Log-log linear regression model with GDP by region interaction"
  ) |>
  # add caption
  tab_caption(
    caption = "Table summarizing the relationship between GDP per capita and epidemiologically linked measles cases using yearly country observations from 2012–2024."
  ) |>
  # bold column labels
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )
```

The model reveals significant trends. The baseline region in this model is Africa. In Africa, there isn't enough statistically significant evidence of a negative association between GDP per capita and epidemiologically linked measles cases. Eastern Mediterranean (EMRO) and South-East Asia (SEARO) are positively and significantly associated with higher epidemiologically linked measles cases than Africa, accounting for GDP level. Western Pacific (WPRO) is negatively and significantly associated with lower measles cases than Africa, accounting for GDP level. The Americas (AMRO) and Europe (EURO) do not show statistically significant differences from Africa at baseline GDP levels.

The interaction terms show that the relationship between GDP per capita and epidemiologically linked measles cases differs by region. In Eastern Mediterranean and South-East Asia, the association between GDP and measles cases is more negative than in Africa, meaning an increase in GDP is associated with a decrease in cases. Contrary to Western Pacific which shows a positive interaction, meaning that the relationship between GDP and measles cases is less negative than in Africa. The interaction effects for the Americas and Europe do not show statistically significant evidence that the relationship between GDP and measles cases differs from Africa in these regions.

Overall, the results demonstrate that GDP per capita's relationship with epidemiologically linked measles cases varies across regions, and that economic development affects disease patterns differently based on geographic context. Specifically, countries in the Eastern Mediterranean and South-East Asia regions with lower GDP per capita were found to be associated with higher numbers of epidemiologically linked measles cases.
